---
title: "DATS610 Final Project- Analyzing trends in U.S. Traffic Accident Data"
author: "Junran Cao, Jordan Dawes, and Carter Rogers"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: true
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=F}
library(dplyr)
library(ggplot2)
library(tidyr)
library(gridExtra)
library(tidyverse)  
library(corrplot)
library(ezids)
library(httr)
library(randomForest)
library(tree)
library(caret)
library(broom)
```


```{r setup, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
# knitr::opts_chunk$set(include = F)
# knitr::opts_chunk$set(echo = TRUE)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

# Background
## Overview

The topic we have chosen to examine in this project is traffic accidents in the United States. According to the National Highway and Transportation Security Agency, over 30,000 fatal car accidents occurred in 2019 alone. The United States is a nation that relies heavily on automobile traffic in order to transport people from point A to point B. Car accidents are an unfortunate reality of our modern transportation infrastructure. The severity of these accidents can range from minor fender-benders to deadly collisions. Our goal for this project is to study the makeup of these accidents to gain potential insights from the data available to us. 

## Prior Research



## Dataset Description
 
 
 
## Limitations of the dataset

There are several limitations of the data set that will complicate our modeling and testing processes. 

## SMART Questions and Hypotheses

1. Can we classify the severity of a given accident based on temperature, wind speed, or other weather conditions present during the accident?

From prior knowledge we can infer that poor weather conditions lead to more severe traffic accidents. Low temperatures and high wind speeds can cause poor conditions that make driving very difficult. 

2. What factors best predict the severity of an accident?

We are curious to see which factors present during an accident best predict the severity of the ensuing accident. We plan to utilize as many variables as possible in the data set to compare their potential relationships with the severity of an accident.

3. Is there a relationship between civil twilight and accident occurrences? Is there any relationship between visibility and the severity of an accident?



4. Is there any relationship between the severity of an accident and the road conditions (traffic stops, railways, junctions, etc.)?



# Analysis 
## Exploratory Data Analysis (EDA)



## Tests and Modeling



## Findings



# Conclusion



# References


